---
title: "alfak2 vs alfakR Real Sample Landscape Comparison"
subtitle: "MINOBS-stratified comparison against legacy nn_prior_empirical_two_shell outputs"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
params:
  patient_subset: null
  min_total_count_grid: [5, 10, 20]
  alfak2_fit_root: "/Users/4482173/Documents/GitHub/alfak2/benchmark/results/real_samples"
  alfakR_fit_root: "/Users/4482173/Documents/GitHub/alfakR/benchmark/results/fits/nn_prior_empirical_two_shell/pm_0.00005"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11,
  fig.height = 5.5,
  dpi = 150
)

repo_dir <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), "../.."), winslash = "/")
source(file.path(repo_dir, "benchmark", "R", "real_sample_helpers.R"))
source(file.path(repo_dir, "benchmark", "R", "legacy_comparison_helpers.R"))
dirs <- ensure_benchmark_dirs(repo_dir, "legacy_comparison")
```

## Comparison Scope

This report compares existing result files. It does not refit either method.

- New method: current `alfak2` real-sample fits from `params$alfak2_fit_root`.
- Legacy method: `alfakR` `nn_prior_empirical_two_shell`, `pm = 0.00005`,
  from `params$alfakR_fit_root`.
- Minimum-observation thresholds are compared separately using
  `params$min_total_count_grid`.

Real samples do not provide a complete truth landscape. Therefore, this report
does not claim absolute accuracy improvement. It reports comparable output
properties: landscape size, overlap on common karyotypes, rank agreement, old
cross-validation R-squared where available, current optimization diagnostics,
and one-step beneficial event differences.

## Run Comparison

```{r run-comparison}
comparison <- run_alfakR_alfak2_real_comparison_grid(
  repo_dir = repo_dir,
  new_fit_root = params$alfak2_fit_root,
  old_fit_root = params$alfakR_fit_root,
  min_total_count_grid = params$min_total_count_grid,
  patient_subset = params$patient_subset
)

comparison$minobs_grid
comparison$landscape_metrics
comparison$beneficial_summary
```

## Landscape Node Counts

```{r node-counts}
p_nodes <- plot_node_count_comparison(comparison$landscape_metrics)
ggplot2::ggsave(file.path(dirs$figures, "alfakR_vs_alfak2_node_counts_by_minobs.png"),
                p_nodes, width = 11, height = 8, dpi = 150)
p_nodes
```

## Common-Karyotype Rank Agreement

```{r common-rank}
p_rank <- plot_common_rank_comparison(comparison$landscape_metrics)
ggplot2::ggsave(file.path(dirs$figures, "alfakR_vs_alfak2_common_rank_spearman_by_minobs.png"),
                p_rank, width = 11, height = 8, dpi = 150)
p_rank
```

## Held-Out Validation

```{r xval}
p_xval <- plot_xval_comparison(comparison$landscape_metrics)
ggplot2::ggsave(file.path(dirs$figures, "alfakR_vs_alfak2_xval_r2_by_minobs.png"),
                p_xval, width = 11, height = 8, dpi = 150)
p_xval
```

## One-Step Beneficial Differences

```{r beneficial-delta}
p_beneficial_delta <- plot_beneficial_delta_heatmap(comparison$beneficial_profile)
ggplot2::ggsave(file.path(dirs$figures, "alfakR_vs_alfak2_beneficial_delta_heatmap_by_minobs.png"),
                p_beneficial_delta, width = 11, height = 9, dpi = 150)
p_beneficial_delta
```

## Output Files

```{r output-files}
list.files(dirs$tables, full.names = TRUE)
list.files(dirs$figures, full.names = TRUE)
file.path(dirs$results, "alfakR_vs_alfak2_real_comparison_by_minobs.rds")
```
