---
title: "alfak2 Controlled Synthetic Landscape Sparsity Benchmark"
subtitle: "Known truth landscape, downsampled sparse inputs, and recovery accuracy"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
params:
  families: ["structured_epistatic", "rugged"]
  pm_grid: [1e-04, 5e-04, 1e-03]
  min_obs_grid: [1]
  downsample_fraction_grid: [0.0002, 0.0005, 0.001]
  replicates: 2
  n_chr: 6
  min_cn: 1
  max_cn: 8
  K: 1000000000
  N0: 100
  step_size: 0.2
  dropout_prob: 0
  seed: 101
  fit_max_nodes: 15000
  local_shell_depth: 0
  global_extra_shell: 1
  force_rebuild_reference: false
  parallel_workers: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  dpi = 150
)

repo_dir <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), "../.."), winslash = "/")
source(file.path(repo_dir, "benchmark", "R", "real_sample_helpers.R"))
source(file.path(repo_dir, "benchmark", "R", "reference_logistic_model.R"))
source(file.path(repo_dir, "benchmark", "R", "synthetic_benchmark_helpers.R"))
load_alfak2_for_benchmark(repo_dir)
dirs <- ensure_benchmark_dirs(repo_dir, "synthetic_sparsity")
```

## Benchmark Design

This benchmark has an explicit truth table and a forward sampling model. For
each synthetic landscape family, it:

1. Generates a complete bounded copy-number landscape with known fitness.
2. Starts from the diploid state with fitness fixed to 1.
3. Evolves the population under logistic growth with carrying capacity `K`.
4. Applies paired gain/loss missegregation events with rate `pm` per chromosome
   per division.
5. Records the original population compositions at `T` and `T + dt`, where
   `T` is the diploid baseline half-capacity time rounded to the nearest
   positive integer and `dt` is `0.8 * T_raw` rounded to the nearest positive
   integer.
6. Caches that standard two-timepoint input under `benchmark/data/` so expensive
   reference generation is skipped on later runs with the same parameters.
7. Downsamples directly from the original population distributions at those
   fixed timepoints, not from a previously sampled count table.
8. Applies `min_obs` rare-node filtering and optional dropout.
9. Fits `alfak2`.
10. Scores posterior landscapes against the known truth.

Reference-standard tasks are parallelizable over `family x replicate x pm`.
Each worker then runs the requested `min_obs x downsample_fraction` grid for
that reference input.

The primary plot is sparsity versus recovery accuracy.

The complete reference landscape size is `(max_cn - min_cn + 1)^n_chr`.
The default is intentionally small enough to enumerate exactly. A full
22-chromosome lattice with copy numbers 1 through 8 would contain
`8^22 = 73,786,976,294,838,206,464` states and is not enumerable or storable
as a dense truth table.

```{r landscape-scale}
data.frame(
  n_chr = params$n_chr,
  min_cn = params$min_cn,
  max_cn = params$max_cn,
  states_per_chr = params$max_cn - params$min_cn + 1,
  complete_states = (params$max_cn - params$min_cn + 1)^params$n_chr
)
```

## Run Benchmark

```{r run-benchmark}
bench <- run_synthetic_sparsity_benchmark(
  repo_dir = repo_dir,
  families = params$families,
  pm_grid = as.numeric(params$pm_grid),
  min_obs_grid = as.integer(params$min_obs_grid),
  downsample_fraction_grid = as.numeric(params$downsample_fraction_grid),
  replicates = as.integer(params$replicates),
  n_chr = as.integer(params$n_chr),
  min_cn = as.integer(params$min_cn),
  max_cn = as.integer(params$max_cn),
  K = as.numeric(params$K),
  N0 = as.numeric(params$N0),
  step_size = as.numeric(params$step_size),
  dropout_prob = as.numeric(params$dropout_prob),
  seed = as.integer(params$seed),
  fit_max_nodes = as.integer(params$fit_max_nodes),
  local_shell_depth = as.integer(params$local_shell_depth),
  global_extra_shell = as.integer(params$global_extra_shell),
  force_rebuild_reference = isTRUE(params$force_rebuild_reference),
  parallel_workers = as.integer(params$parallel_workers)
)

bench$reference_summary
bench$metrics
bench$elbow
```

## Sparsity vs Correlation

```{r plot-spearman}
p_spearman <- plot_sparsity_performance(bench$metrics, metric = "spearman", x = "downsample_fraction")
ggplot2::ggsave(file.path(dirs$figures, "synthetic_sparsity_vs_spearman.png"),
                p_spearman, width = 10, height = 5.5, dpi = 150)
p_spearman
```

```{r plot-pearson}
p_pearson <- plot_sparsity_performance(bench$metrics, metric = "pearson", x = "downsample_fraction")
ggplot2::ggsave(file.path(dirs$figures, "synthetic_sparsity_vs_pearson.png"),
                p_pearson, width = 10, height = 5.5, dpi = 150)
p_pearson
```

## Sparsity vs Error

```{r plot-rmse}
p_rmse <- plot_error_vs_depth(bench$metrics, metric = "rmse")
ggplot2::ggsave(file.path(dirs$figures, "synthetic_depth_vs_rmse.png"),
                p_rmse, width = 10, height = 5.5, dpi = 150)
p_rmse
```

```{r plot-mae}
p_mae <- plot_error_vs_depth(bench$metrics, metric = "mae")
ggplot2::ggsave(file.path(dirs$figures, "synthetic_depth_vs_mae.png"),
                p_mae, width = 10, height = 5.5, dpi = 150)
p_mae
```

## Calibration and Top-K Recovery

```{r calibration}
calibration_tbl <- aggregate(
  cbind(coverage95, mean_width95, topk_overlap) ~ family + pm + min_obs + layer + downsample_fraction,
  data = bench$metrics,
  FUN = mean,
  na.rm = TRUE
)
utils::write.csv(calibration_tbl, file.path(dirs$tables, "synthetic_sparsity_calibration.csv"), row.names = FALSE)
calibration_tbl
```

## Output Files

```{r output-files}
list.files(dirs$tables, full.names = TRUE)
list.files(dirs$figures, full.names = TRUE)
```
