---
title: "alfak2 Real Sample Landscapes"
subtitle: "Run real two-timepoint samples and summarize one-step beneficial CN events"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
params:
  patient_subset: ["P2"]
  min_total_count: 20
  drop_diploid: true
  beta: 0.01
  min_cn: 0
  max_cn: 5
  local_shell_depth: 0
  global_extra_shell: 1
  max_nodes: 4000
  force_refit: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11,
  fig.height = 5.5,
  dpi = 150
)

repo_dir <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), "../.."), winslash = "/")
source(file.path(repo_dir, "benchmark", "R", "real_sample_helpers.R"))
load_alfak2_for_benchmark(repo_dir)
dirs <- ensure_benchmark_dirs(repo_dir, "real_samples")
```

## Input Samples

The real-sample workflow reads local, non-versioned perspective data from
`benchmark/data`. It builds a two-column karyotype count matrix for each
patient, fits `alfak2`, saves the posterior landscape, and then summarizes
one-step copy-number moves from directly informed states.

The default settings keep the local TMB fit on directly observed nodes and add
one global shell for graph-borrowed one-step neighbors. Set
`local_shell_depth: 1` if you want the local hierarchical layer itself to
include one-step borrowed nodes; that is slower for 22-chromosome real samples.

```{r input-patients}
data_dir <- file.path(repo_dir, "benchmark", "data")
available_real_patient_ids(data_dir)
```

## Run alfak2

```{r fit-real-samples}
bundle_list <- run_real_sample_landscapes(
  repo_dir = repo_dir,
  patient_subset = params$patient_subset,
  min_total_count = params$min_total_count,
  drop_diploid = isTRUE(params$drop_diploid),
  beta = params$beta,
  min_cn = params$min_cn,
  max_cn = params$max_cn,
  local_shell_depth = params$local_shell_depth,
  global_extra_shell = params$global_extra_shell,
  max_nodes = params$max_nodes,
  force = isTRUE(params$force_refit)
)

length(bundle_list)
names(bundle_list)
```

```{r fit-summary}
fit_summary <- do.call(rbind, lapply(bundle_list, function(x) {
  data.frame(
    patient_id = x$patient_id,
    observed_karyotypes = nrow(x$input$counts),
    t0_cells = sum(x$input$counts[, 1]),
    t1_cells = sum(x$input$counts[, 2]),
    dt = x$input$dt,
    local_nodes = nrow(x$fit$local$summary),
    global_nodes = nrow(x$fit$global$summary),
    convergence = x$fit$local$diagnostics$convergence,
    gradient_norm = x$fit$local$diagnostics$gradient_norm,
    stringsAsFactors = FALSE
  )
}))

utils::write.csv(fit_summary, file.path(dirs$tables, "real_sample_fit_summary.csv"), row.names = FALSE)
fit_summary
```

## Landscape Distribution

```{r landscape-distribution}
p_landscape <- plot_landscape_distribution(bundle_list)
ggplot2::ggsave(file.path(dirs$figures, "real_sample_landscape_distribution.png"),
                p_landscape, width = 11, height = 5.5, dpi = 150)
p_landscape
```

## One-Step Beneficial Matrix

For each directly informed parent state, each available one-step gain/loss
neighbor is evaluated as:

`delta = fitness(child) - fitness(parent)`

The matrix below shows the proportion of valid one-step moves that are
beneficial for each patient and chromosome event.

```{r beneficial-matrices}
beneficial <- beneficial_matrices_from_bundles(bundle_list)

write_matrix_csv(
  beneficial$proportion,
  file.path(dirs$tables, "real_sample_one_step_beneficial_proportion_matrix.csv")
)
write_matrix_csv(
  beneficial$valid_n,
  file.path(dirs$tables, "real_sample_one_step_valid_n_matrix.csv")
)
utils::write.csv(
  beneficial$edges,
  file.path(dirs$tables, "real_sample_one_step_edges.csv"),
  row.names = FALSE
)

beneficial$proportion[, seq_len(min(10, ncol(beneficial$proportion))), drop = FALSE]
```

```{r beneficial-heatmap}
p_beneficial <- plot_beneficial_matrix(beneficial$proportion)
ggplot2::ggsave(file.path(dirs$figures, "real_sample_one_step_beneficial_heatmap.png"),
                p_beneficial, width = 11, height = 5.5, dpi = 150)
p_beneficial
```

```{r valid-edge-heatmap}
p_valid <- plot_valid_matrix(beneficial$valid_n)
ggplot2::ggsave(file.path(dirs$figures, "real_sample_one_step_valid_edges_heatmap.png"),
                p_valid, width = 11, height = 5.5, dpi = 150)
p_valid
```

## Top One-Step Fitness Gains

```{r top-beneficial-edges}
top_edges <- beneficial$edges[is.finite(beneficial$edges$delta), , drop = FALSE]
top_edges <- top_edges[order(top_edges$patient_id, -top_edges$delta), , drop = FALSE]
top_edges <- do.call(rbind, lapply(split(top_edges, top_edges$patient_id), head, 15))
utils::write.csv(top_edges, file.path(dirs$tables, "real_sample_top_one_step_edges.csv"), row.names = FALSE)
top_edges
```
