---
title: "alfak2 Real Sample Landscapes"
subtitle: "Run real two-timepoint samples and summarize one-step beneficial CN events"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
params:
  patient_subset: ["P2"]
  min_total_count_grid: [5, 10, 20]
  drop_diploid: true
  beta: 0.00005
  min_cn: 0
  max_cn: 5
  local_shell_depth: 1
  global_extra_shell: 1
  max_nodes: 15000
  input_depth: "raw"
  effective_depth: null
  effective_depth_mode: "min"
  observation_model: null
  dm_concentration: null
  output_root: "real_samples"
  parallel_workers: Inf
  compute_xval: true
  xval_max_folds: Inf
  xval_seed: 1
  force_refit: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11,
  fig.height = 5.5,
  dpi = 150
)

repo_dir <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), "../.."), winslash = "/")
source(file.path(repo_dir, "benchmark", "R", "real_sample_helpers.R"))
load_alfak2_for_benchmark(repo_dir)
dirs <- ensure_benchmark_dirs(repo_dir, "real_samples")
```

## Input Samples

The real-sample workflow reads local, non-versioned perspective data from
`benchmark/data`. It builds a two-column karyotype count matrix for each
patient and each requested minimum-observation threshold, fits `alfak2`, saves
the posterior landscape, and then summarizes one-step copy-number moves from
directly informed states.

The default settings fit directly observed nodes plus one-step neighbors in the
local TMB layer, then add one global shell for graph-borrowed two-step
neighbors. Each
`patient_id x min_total_count` task is scheduled in parallel, while each
individual `alfak2` fit is run serially. `parallel_workers: Inf` means one
worker per task; lower it to a fixed integer if memory pressure is high.

```{r input-patients}
data_dir <- file.path(repo_dir, "benchmark", "data")
available_real_patient_ids(data_dir)
```

## Run alfak2

```{r fit-real-samples}
bundle_list <- run_real_sample_landscape_grid(
  repo_dir = repo_dir,
  patient_subset = params$patient_subset,
  min_total_count_grid = params$min_total_count_grid,
  drop_diploid = isTRUE(params$drop_diploid),
  beta = params$beta,
  min_cn = params$min_cn,
  max_cn = params$max_cn,
  local_shell_depth = params$local_shell_depth,
  global_extra_shell = params$global_extra_shell,
  max_nodes = params$max_nodes,
  input_depth = params$input_depth,
  effective_depth = if (is.null(params$effective_depth)) NULL else as.numeric(params$effective_depth),
  effective_depth_mode = params$effective_depth_mode,
  observation_model = if (is.null(params$observation_model)) NULL else params$observation_model,
  dm_concentration = if (is.null(params$dm_concentration)) NULL else as.numeric(params$dm_concentration),
  output_root = params$output_root,
  parallel_workers = params$parallel_workers,
  compute_xval = isTRUE(params$compute_xval),
  xval_max_folds = as.numeric(params$xval_max_folds),
  xval_seed = as.integer(params$xval_seed),
  force = isTRUE(params$force_refit)
)

if (!length(bundle_list)) {
  stop("No successful alfak2 tasks were produced.", call. = FALSE)
}

bundle_by_minobs <- split(
  bundle_list,
  vapply(bundle_list, function(x) as.character(x$min_total_count), character(1))
)
bundle_by_minobs <- lapply(bundle_by_minobs, function(x) {
  names(x) <- vapply(x, function(y) y$patient_id, character(1))
  x
})

length(bundle_list)
names(bundle_list)
```

```{r fit-summary}
fit_summary <- do.call(rbind, lapply(bundle_list, function(x) {
  task_id <- if (!is.null(x$task_id)) {
    x$task_id
  } else {
    paste0(x$patient_id, "_MINOBS_", x$min_total_count)
  }
  data.frame(
    task_id = task_id,
    patient_id = x$patient_id,
    min_total_count = x$min_total_count,
    output_subdir = x$output_subdir,
    local_shell_depth = x$fit_parameters$local_shell_depth,
    global_extra_shell = x$fit_parameters$global_extra_shell,
    observed_karyotypes = nrow(x$input$counts),
    t0_cells = sum(x$input$counts[, 1]),
    t1_cells = sum(x$input$counts[, 2]),
    dt = x$input$dt,
    local_nodes = nrow(x$fit$local$summary),
    global_nodes = nrow(x$fit$global$summary),
    xval_r2 = if (!is.null(x$xval$R2R)) x$xval$R2R else NA_real_,
    xval_n = if (!is.null(x$xval$detail)) nrow(x$xval$detail) else NA_integer_,
    convergence = x$fit$local$diagnostics$convergence,
    gradient_norm = x$fit$local$diagnostics$gradient_norm,
    stringsAsFactors = FALSE
  )
}))
fit_summary <- fit_summary[order(fit_summary$min_total_count, fit_summary$patient_id), , drop = FALSE]

utils::write.csv(fit_summary, file.path(dirs$tables, "real_sample_fit_summary_by_minobs.csv"), row.names = FALSE)
utils::write.csv(fit_summary, file.path(dirs$tables, "real_sample_fit_summary.csv"), row.names = FALSE)
fit_summary
```

## Landscape Distribution

```{r landscape-distribution}
landscape_plots <- lapply(names(bundle_by_minobs), function(min_obs) {
  suffix <- paste0("MINOBS_", min_obs)
  p <- plot_landscape_distribution(bundle_by_minobs[[min_obs]]) +
    ggplot2::ggtitle(paste("Landscape distribution", suffix))
  ggplot2::ggsave(
    file.path(dirs$figures, paste0("real_sample_landscape_distribution_", suffix, ".png")),
    p,
    width = 11,
    height = 5.5,
    dpi = 150
  )
  print(p)
  p
})
names(landscape_plots) <- names(bundle_by_minobs)
```

## One-Step Beneficial Matrix

For each directly informed parent state, each available one-step gain/loss
neighbor is evaluated as:

`delta = fitness(child) - fitness(parent)`

The matrix below shows the proportion of valid one-step moves that are
beneficial for each patient and chromosome event.

```{r beneficial-matrices}
beneficial_by_minobs <- lapply(names(bundle_by_minobs), function(min_obs) {
  suffix <- paste0("MINOBS_", min_obs)
  beneficial <- beneficial_matrices_from_bundles(bundle_by_minobs[[min_obs]])
  write_matrix_csv(
    beneficial$proportion,
    file.path(dirs$tables, paste0("real_sample_one_step_beneficial_proportion_matrix_", suffix, ".csv"))
  )
  write_matrix_csv(
    beneficial$valid_n,
    file.path(dirs$tables, paste0("real_sample_one_step_valid_n_matrix_", suffix, ".csv"))
  )
  utils::write.csv(
    beneficial$edges,
    file.path(dirs$tables, paste0("real_sample_one_step_edges_", suffix, ".csv")),
    row.names = FALSE
  )
  beneficial
})
names(beneficial_by_minobs) <- names(bundle_by_minobs)

beneficial_by_minobs[[1]]$proportion[, seq_len(min(10, ncol(beneficial_by_minobs[[1]]$proportion))), drop = FALSE]
```

```{r beneficial-heatmap}
beneficial_plots <- lapply(names(beneficial_by_minobs), function(min_obs) {
  suffix <- paste0("MINOBS_", min_obs)
  p <- plot_beneficial_matrix(beneficial_by_minobs[[min_obs]]$proportion) +
    ggplot2::ggtitle(paste("One-step beneficial proportion", suffix))
  ggplot2::ggsave(
    file.path(dirs$figures, paste0("real_sample_one_step_beneficial_heatmap_", suffix, ".png")),
    p,
    width = 11,
    height = 5.5,
    dpi = 150
  )
  print(p)
  p
})
names(beneficial_plots) <- names(beneficial_by_minobs)
```

```{r valid-edge-heatmap}
valid_plots <- lapply(names(beneficial_by_minobs), function(min_obs) {
  suffix <- paste0("MINOBS_", min_obs)
  p <- plot_valid_matrix(beneficial_by_minobs[[min_obs]]$valid_n) +
    ggplot2::ggtitle(paste("Valid one-step edges", suffix))
  ggplot2::ggsave(
    file.path(dirs$figures, paste0("real_sample_one_step_valid_edges_heatmap_", suffix, ".png")),
    p,
    width = 11,
    height = 5.5,
    dpi = 150
  )
  print(p)
  p
})
names(valid_plots) <- names(beneficial_by_minobs)
```

## Top One-Step Fitness Gains

```{r top-beneficial-edges}
top_edges <- do.call(rbind, lapply(names(beneficial_by_minobs), function(min_obs) {
  suffix <- paste0("MINOBS_", min_obs)
  edges <- beneficial_by_minobs[[min_obs]]$edges
  edges <- edges[is.finite(edges$delta), , drop = FALSE]
  edges <- edges[order(edges$patient_id, -edges$delta), , drop = FALSE]
  edges <- if (nrow(edges)) {
    do.call(rbind, lapply(split(edges, edges$patient_id), head, 15))
  } else {
    data.frame()
  }
  if (!is.null(edges) && nrow(edges)) {
    edges$min_total_count <- as.integer(min_obs)
    utils::write.csv(
      edges,
      file.path(dirs$tables, paste0("real_sample_top_one_step_edges_", suffix, ".csv")),
      row.names = FALSE
    )
  }
  edges
}))
if (is.null(top_edges)) top_edges <- data.frame()
utils::write.csv(
  top_edges,
  file.path(dirs$tables, "real_sample_top_one_step_edges_by_minobs.csv"),
  row.names = FALSE
)
top_edges
```
